---
title: "Orthogroup visualisation - overview statistics"
author: "Sofie Agerbæk"
date: "2025-10-13"
output: html_document
---

```{r, include=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ComplexUpset)
```


```{r, include=FALSE}
library(cowplot)
options(digits = 3)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 3) # number of digits printed by tibbles default.
options(width = 200)

text_base_size   <- 9 # in pt
fig.witdh        <- 180  # in mm
fig.height       <- 125  # in mm

# Set all text in plots to same size
theme_set(theme_cowplot(font_size = 10, rel_small = 1, rel_tiny = 1, rel_large = 1))
theme_update(panel.grid.major = element_line(colour="gray85"), panel.grid.minor = element_line(colour="gray95", linetype = 7))
# Setting output sizes for plots
knitr::opts_chunk$set(fig.width = fig.witdh/25.4)
knitr::opts_chunk$set(fig.height = fig.height/25.4)
knitr::opts_chunk$set(dpi = 108) # You need to find your monitors dpi yourself.

# Setting text size inside plots (geom_text, geom_label etc.)
ggplot_text_size <- text_base_size / ggplot2::.pt
# Now use: geom_text(..., size = ggplot_text_size)


options(ggplot2.continuous.fill  = scale_fill_viridis_c)
options(ggplot2.discrete.fill    = c("slategray2", "#404e67","#809bce","#9887D9" ,"#9C5BA2", "#DB9FF6",  "#d1625c", "#ffd6a5", "#e89c81"))
options(ggplot2.continuous.colour = scale_colour_viridis_c)
options(ggplot2.discrete.colour   = c("slategray2","#404e67", "#809bce","#9C5BA2","#9887D9",  "#d1625c", "#ffd6a5", "#e89c81"))

pall1_light <- c("#979797","#A6AAB1", "#869BBE", "#6585BD", "#76A1F1","#9DC8FF", "#A3E1EF", "#96FBEA", "#8CDBD1",  "#84D09A", "#7DCF7E", "#B0EE86", "#CEF7A5", "#DCFFC2", "#F2FFA8","#F4FF7B", "#F2F853", "#DDDB63", "#C7BB4E", "#BF9E43", "#CB883E", "#E78332", "#EA6C3D", "#DD5A3D", "#E44242", "#D65050",  "#E27373", "#F27D7C", "#F99A99", "#FFB7B5", "#E990A1","#D1768E" ,"#C9799F", "#BF89B1", "#B599C4","#ABA9D6",  "#987EBD", "#9B8EAC")

```

##### Species order
```{r}
species_order <- c("Rhodotorula toruloides", "Cryptococcus neoformans", "Trichosporon ovoides",
                   "Malassezia furfur", "Malassezia equina", "Malassezia dermatis", "Malassezia sympodialis",
                   "Malassezia pachydermatis", "Malassezia restricta", "Malassezia globosa",
                   "Pichia kudriavzevii", "Saccharomyces cerevisiae", "Nakaseomyces glabratus",
                   "Candida metapsilosis", "Candida orthopsilosis", "Candida parapsilosis", "Candida tropicalis",
                   "Candida albicans", "Candida dubliniensis",
                   "Hortaea werneckii",  "Aspergillus fumigatus", "Histoplasma capsulatum",
                   "Chrysosporium queenslandicum",
                   "Microsporum canis", "Microsporum audouinii", "Microsporum ferrugineum",
                   "Nannizzia gypsea", "Epidermophyton floccosum", "Trichophyton schoenleinii",
                   "Trichophyton tonsurans", "Trichophyton mentagrophytes", "Trichophyton indotineae",
                   "Trichophyton interdigitale", "Trichophyton concentricum", "Trichophyton soudanese",
                   "Trichophyton kuryangei", "Trichophyton rubrum", "Trichophyton violaceum" )

```


## Overall statistics

Reading the statistics file
```{r}
Species_stats <- read.csv("../results/Results_Aug23/Comparative_Genomics_Statistics/Statistics_PerSpecies.tsv", sep = "\t", na.strings = "")

stat_df <- head(Species_stats, 10)
rownames(stat_df) <- stat_df$X
stat_df$X <- NULL
stat_df <- data.frame(t(stat_df))

rm(Species_stats)

head(stat_df[1:3])

stat_df <- rownames_to_column(stat_df, var = "sample_id")
stat_df$sample_id <- sub("\\.", "_", stat_df$sample_id)

extracted_strings <- lapply(strsplit((stat_df$sample_id), "_"), function(x) paste(x[1:2], collapse = "_"))

stat_df$species <- unlist(extracted_strings)

stat_df[, 2:11] <- apply(stat_df[, 2:11], 2, function(x) as.numeric(as.character(x)))

```


Plotting the number of genes for each sample along with the number of those genes in orthogroups
```{r, fig.width=6, fig.height=9}
k <- ggplot(stat_df, aes(y = sample_id))+
  geom_col(aes(x=Number.of.genes, fill = as.factor(species) ))+
  geom_col(aes(x=Number.of.genes.in.orthogroups), alpha = 0.8,
           width = 0.2)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_manual(values = as.vector(pall1_light))+
  guides(fill = 'none')

k

```

We observe comparable gene numbers within closely related samples.
As expected we see generally smaller amounts of genes in orthogroups among the 'outgroups'.

```{r}
rownames(stat_df) <- stat_df$sample_id
print((stat_df[4:5]))
```

### Species overlap

Looking at gene counts - ie. number of genes in each OG for each sample
```{r}
OG_sp_overlaps <- read.csv("../results/Results_Aug23/Comparative_Genomics_Statistics/Orthogroups_SpeciesOverlaps.tsv",
                          header = TRUE, sep = "\t")

OG_sp_overlaps$X <- sub("\\.", "_", OG_sp_overlaps$X)
OG_sp_overlaps   <-column_to_rownames(OG_sp_overlaps, var = "X")

split_name_col <- strsplit(rownames(OG_sp_overlaps), "_")

genus    <- sapply(split_name_col, `[`, 1)

metadata <- data.frame(
  genus = genus,
  row.names = rownames(OG_sp_overlaps)
)


```

```{r}
library(pheatmap)

mat <- as.matrix(OG_sp_overlaps)

mat[lower.tri(mat)] <- t(mat)[lower.tri(mat)]

mat_pct <- sweep(mat, 2, apply(mat, 2, max), "/") * 100


# Ensure metadata has a column 'Family' and rownames match mat_pct
metadata <- metadata[rownames(mat_pct), , drop = FALSE]


# plot
pheatmap(
  mat_pct,
  annotation_row = metadata,
  annotation_col = metadata,
  color = colorRampPalette(c("white", "gold", "red3"))(100),
  border_color = NA,
  fontsize = 10,
  main = "Orthogroup Overlap Between Species",
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  show_rownames = FALSE
)

```

```{r}
rm(mat, extracted_strings, mat_pct, metadata, OG_sp_overlaps, split_name_col, stat_df, k, genus)
```


## Distribution of orthogroups across samples (figure 3)

For each sample we want to define family, genus and species

```{r}
N0 <- read.csv("../results/Results_Aug23/Phylogenetic_Hierarchical_Orthogroups/N0.tsv", sep = "\t", header = TRUE)
head(N0[1:5])
```


We want just presence/absence in each OG across all species
```{r}
OG_count_df <- N0 %>%
            select(-c("HOG", "Gene.Tree.Parent.Clade")) %>%
            mutate(across(2:67, ~ . != ""))  %>% #if the cell is not empty (ie the species has a gene in that OG)
            group_by(OG) %>%
            summarise(across(everything(), ~ any(.x == TRUE)), .groups = "drop")  #collapsing the multiple rows pr OG (which exist bc of the HOG that we just removed)

names(OG_count_df) <- sub("\\.", "_", names(OG_count_df))
```

Taxa look-up data frame
```{r}
taxonomy <- data.frame(
  sample = c(
    "Aspergillus_fumigatus_IP24",
    "Candida_albicans_F133_05",
    "Candida_dubliniensis_11945",
    "Candida_dubliniensis_2747",
    "Candida_dubliniensis_8501",
    "Candida_metapsilosis_Y_48470",
    "Candida_orthopsilosis_Y_48468",
    "Candida_parapsilosis_11301",
    "Candida_parapsilosis_1954",
    "Candida_tropicalis_MYCT_0816",
    "Chrysosporium_queenslandicum_28077",
    "Cryptococcus_neoformans_KN99",
    "Epidermophyton_floccosum_100148",
    "Epidermophyton_floccosum_10867",
    "Epidermophyton_floccosum_130802",
    "Histoplasma_capsulatum_G186A",
    "Hortaea_werneckii_100455",
    "Hortaea_werneckii_1163",
    "Hortaea_werneckii_41051",
    "Malassezia_dermatis_9169",
    "Malassezia_equina_9969",
    "Malassezia_furfur_141402",
    "Malassezia_furfur_1878",
    "Malassezia_furfur_6000",
    "Malassezia_globosa_7705",
    "Malassezia_globosa_7886",
    "Malassezia_globosa_8744",
    "Malassezia_pachydermatis_21004",
    "Malassezia_restricta_7877",
    "Malassezia_restricta_7991",
    "Malassezia_sympodialis_44340",
    "Microsporum_audouinii_119449",
    "Microsporum_audouinii_33268",
    "Microsporum_canis_113480",
    "Microsporum_canis_15669",
    "Microsporum_canis_44551",
    "Microsporum_ferrugineum_131111",
    "Microsporum_ferrugineum_37371",
    "Nakaseomyces_glabratus_11311",
    "Nakaseomyces_glabratus_138",
    "Nakaseomyces_glabratus_15698",
    "Nannizzia_gypsea_17164",
    "Pichia_kudriavzevii_5732",
    "Rhodotorula_toruloides_Z1",
    "Saccharomyces_cerevisiae_wine010",
    "Trichophyton_concentricum_109405",
    "Trichophyton_concentricum_19626",
    "Trichophyton_indotineae_BE17",
    "Trichophyton_interdigitale_BE4",
    "Trichophyton_kuryangei_IHEM13979",
    "Trichophyton_mentagrophytes_124404",
    "Trichophyton_mentagrophytes_43573",
    "Trichophyton_rubrum_117539",
    "Trichophyton_rubrum_17065",
    "Trichophyton_rubrum_28986",
    "Trichophyton_rubrum_73588",
    "Trichophyton_schoenleinii_45859",
    "Trichophyton_schoenleinii_85571",
    "Trichophyton_soudanese_IHEM19743",
    "Trichophyton_tonsurans_112188",
    "Trichophyton_tonsurans_12935",
    "Trichophyton_tonsurans_72988",
    "Trichophyton_violaceum_73088",
    "Trichosporon_ovoides_4098",
    "Trichosporon_ovoides_7612",
    "Trichosporon_ovoides_9430"
  ),
  species = c(
    "Aspergillus_fumigatus" ,"Candida_albicans_F133","Candida_dubliniensis"
    ,"Candida_dubliniensis", "Candida_dubliniensis","Candida_metapsilosis_Y",
    "Candida_orthopsilosis_Y","Candida_parapsilosis", "Candida_parapsilosis","Candida_tropicalis_MYCT","Chrysosporium_queenslandicum",
    "Cryptococcus_neoformans","Epidermophyton_floccosum","Epidermophyton_floccosum",
    "Epidermophyton_floccosum","Histoplasma_capsulatum","Hortaea_werneckii" ,"Hortaea_werneckii","Hortaea_werneckii","Malassezia_dermatis","Malassezia_equina",
    "Malassezia_furfur","Malassezia_furfur","Malassezia_furfur","Malassezia_globosa",
    "Malassezia_globosa","Malassezia_globosa","Malassezia_pachydermatis",
    "Malassezia_restricta","Malassezia_restricta","Malassezia_sympodialis",
    "Microsporum_audouinii","Microsporum_audouinii","Microsporum_canis",
    "Microsporum_canis","Microsporum_canis","Microsporum_ferrugineum",
    "Microsporum_ferrugineum","Nakaseomyces_glabratus","Nakaseomyces_glabratus",
    "Nakaseomyces_glabratus", "Nannizzia_gypsea","Pichia_kudriavzevii",
    "Rhodotorula_toruloides","Saccharomyces_cerevisiae", "Trichophyton_concentricum",
    "Trichophyton_concentricum" ,   "Trichophyton_indotineae"  ,   "Trichophyton_interdigitale",
    "Trichophyton_kuryangei"   ,    "Trichophyton_mentagrophytes" , "Trichophyton_mentagrophytes",
    "Trichophyton_rubrum"      ,    "Trichophyton_rubrum"     ,    "Trichophyton_rubrum"    ,
    "Trichophyton_rubrum"     ,     "Trichophyton_schoenleinii"  ,  "Trichophyton_schoenleinii",
    "Trichophyton_soudanese"     ,  "Trichophyton_tonsurans"   ,   "Trichophyton_tonsurans"  ,
    "Trichophyton_tonsurans"    ,   "Trichophyton_violaceum"  ,     "Trichosporon_ovoides"   ,
    "Trichosporon_ovoides"       ,  "Trichosporon_ovoides"),
  genus = c(
    "Aspergillus",
    rep("Candida", 9),
    "Chrysosporium",
    "Cryptococcus",
    rep("Epidermophyton", 3),
    "Histoplasma",
    rep("Hortaea", 3),
    rep("Malassezia", 12),
    rep("Microsporum", 7),
    rep("Nakaseomyces", 3),
    "Nannizzia",
    "Pichia",
    "Rhodotorula",
    "Saccharomyces",
    rep("Trichophyton", 18),
    rep("Trichosporon", 3)
  ),
  family = c(
    "Aspergillaceae",     # Aspergillus
    rep("Debaryomycetaceae", 9),  # Candida spp.
    "Onygenaceae",        # Chrysosporium
    "Cryptococcaceae",       # Cryptococcus
    rep("Arthrodermataceae", 3),  # Epidermophyton
    "Ajellomycetaceae",   # Histoplasma
    rep("Teratosphaeriaceae", 3), # Hortaea
    rep("Malasseziaceae", 12),    # Malassezia spp.
    rep("Arthrodermataceae", 7),  # Microsporum spp.
    rep("Saccharomycetaceae", 3), # Nakaseomyces glabratus
    "Arthrodermataceae",  # Nannizzia
    "Saccharomycetaceae", # Pichia
    "Sporidiobolaceae",   # Rhodotorula
    "Saccharomycetaceae", # Saccharomyces
    rep("Arthrodermataceae", 18), # Trichophyton spp.
    rep("Trichosporonaceae", 3)   # Trichosporon spp.
  ),
  stringsAsFactors = FALSE
)

```

Merge the OG-counts with the taxonomy of the samples, and then summarise the distribution of OGs in each taxonomic category
```{r}
# Convert to long format
OG_count_long <- OG_count_df %>%
  pivot_longer(cols = 2:67, names_to = "sample", values_to = "present") %>%
  filter(present)

# Add taxonomy info
OG_count_long <- OG_count_long %>%
  left_join(taxonomy, by = "sample")

# For each OG, find which families/genera/species have it
og_groups <- OG_count_long %>%
  group_by(OG) %>%
  summarise(
    n_sample    = n(),
    n_species   = n_distinct(species),
    n_genera    = n_distinct(genus),
    n_families  = n_distinct(family),
    all_species = n_sample == ncol(OG_count_df) - 1
  )

# Assign a category to each OG
og_groups <- og_groups %>%
  mutate(category = case_when(
    all_species ~ "All",
    n_families > 1 ~ "Several families",
    n_families == 1 & n_genera > 1 ~ "Family",
    n_families == 1 & n_genera == 1 & n_species > 1 ~ "Genus",
    n_species == 1 ~ "Species",
    TRUE ~ "Other"
  ))
```

Count OGs in each sample in each category
```{r}
OG_count <- OG_count_long %>%
  left_join(og_groups %>% select(OG, category), by = "OG") %>%
  group_by(sample, category) %>%
  summarise(count = n(), .groups = "drop")

OG_count <- OG_count %>%
  group_by(sample) %>%
  mutate(percent = count / sum(count) * 100)

```

Tree order - so it matches with the phylo-tree for fig 2
```{r}
sample_order <- c( "Rhodotorula_toruloides_Z1", "Cryptococcus_neoformans_KN99", "Trichosporon_ovoides_9430",
                   "Trichosporon_ovoides_7612", "Trichosporon_ovoides_4098",
                   "Malassezia_furfur_141402",  "Malassezia_furfur_1878", "Malassezia_furfur_6000",
                   "Malassezia_equina_9969" , "Malassezia_dermatis_9169", "Malassezia_sympodialis_44340" ,
                   "Malassezia_pachydermatis_21004", "Malassezia_restricta_7991" , "Malassezia_restricta_7877"  ,
                   "Malassezia_globosa_7705"  ,  "Malassezia_globosa_8744" , "Malassezia_globosa_7886",
                   "Pichia_kudriavzevii_5732" ,"Saccharomyces_cerevisiae_wine010","Nakaseomyces_glabratus_138",
                   "Nakaseomyces_glabratus_15698", "Nakaseomyces_glabratus_11311",
                   "Candida_metapsilosis_Y_48470", "Candida_orthopsilosis_Y_48468" ,"Candida_parapsilosis_1954",
                   "Candida_parapsilosis_11301", "Candida_tropicalis_MYCT_0816", "Candida_albicans_F133_05",
                   "Candida_dubliniensis_2747",  "Candida_dubliniensis_11945", "Candida_dubliniensis_8501",
                   "Hortaea_werneckii_1163", "Hortaea_werneckii_41051", "Hortaea_werneckii_100455",
                   "Aspergillus_fumigatus_IP24","Histoplasma_capsulatum_G186A","Chrysosporium_queenslandicum_28077",
                   "Microsporum_canis_44551", "Microsporum_canis_113480", "Microsporum_canis_15669",
                   "Microsporum_audouinii_33268",      "Microsporum_audouinii_119449",
                   "Microsporum_ferrugineum_37371", "Microsporum_ferrugineum_131111",
                   "Nannizzia_gypsea_17164", "Epidermophyton_floccosum_10867", "Epidermophyton_floccosum_130802",
                   "Epidermophyton_floccosum_100148",
                   "Trichophyton_schoenleinii_85571",  "Trichophyton_schoenleinii_45859",
                   "Trichophyton_tonsurans_112188","Trichophyton_tonsurans_12935", "Trichophyton_tonsurans_72988",
                   "Trichophyton_mentagrophytes_124404", "Trichophyton_indotineae_BE17",
                   "Trichophyton_interdigitale_BE4", "Trichophyton_mentagrophytes_43573",
                    "Trichophyton_concentricum_109405", "Trichophyton_concentricum_19626",
                   "Trichophyton_soudanese_IHEM19743",   "Trichophyton_kuryangei_IHEM13979",
                   "Trichophyton_rubrum_17065","Trichophyton_rubrum_117539","Trichophyton_rubrum_28986",
                   "Trichophyton_rubrum_73588",   "Trichophyton_violaceum_73088")

```

### Fig 3a
Plotting barplot
```{r, fig.height=9, fig.width=4}
OG_count$category <- factor(OG_count$category,
                      levels = rev(c("All", "Several families", "Family", "Genus", "Species")))

OG_count$sample <- factor(OG_count$sample,
                      levels = rev(sample_order))

OG_dist_plot <- ggplot(OG_count, aes(y = sample, x = count, fill = category))+
  geom_bar(stat = "identity", width = 0.85) +
  labs(
    title = "Distribution of Orthogroups per Species",
    x = NULL,
    y = NULL,
    fill = "Category"
  ) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        axis.line = element_line(colour = "grey20", linewidth = 0.5), axis.ticks.x = element_blank(),
        axis.ticks.y = element_line(colour = "grey20", linewidth = 0.5))+
  scale_x_continuous(expand = expansion(mult = 0, add = 100))

OG_dist_plot
```


```{r}
# pdf("../plots/OG_dist_plot.pdf", height = 9, width = 6.5)
# OG_dist_plot
# dev.off()
```

```{r}
OG_count_wide <- OG_count %>%
  ungroup() %>%
  select(-percent) %>%
  pivot_wider(names_from = sample, values_from = count)

OG_percent_wide <- OG_count %>%
  mutate(percent = round(percent, digits = 1)) %>%
  ungroup() %>%
  select(-count) %>%
  pivot_wider(names_from = sample, values_from = percent)

# write.csv(OG_count_wide, "../generated/OG_dist_counts.csv")
# write.csv(OG_percent_wide, "../generated/OG_dist_percent.csv")
```


### Fig 3b - upset plot

Upset on genus level
Making upset dataframe with presence/absence for each genus in each OG
```{r}
upset_df <-  OG_count_long %>%
  select(-c("sample", "species", "family")) %>%
  group_by(OG, genus) %>%
  summarise(present = any(present), .groups = "drop") %>%
  pivot_wider(names_from = "genus", values_from = present, values_fill = FALSE)
```

Plotting (takes a little time)
```{r, fig.width=10, fig.height=7}
intersect <- rev(c("Rhodotorula",   "Cryptococcus",  "Trichosporon" ,  "Malassezia" , "Pichia",  "Saccharomyces",  "Nakaseomyces",   "Candida", "Hortaea",  "Aspergillus"   ,   "Histoplasma" ,    "Chrysosporium","Microsporum" , "Nannizzia" , "Epidermophyton", "Trichophyton" ))

upsetplot <- upset(upset_df,
      intersect = intersect,
      n_intersections = 18, height_ratio = c(0.99, 0.01),
      width_ratio = c(0.2, 0.8),
      sort_sets = F,
      matrix=(
        intersection_matrix(
            geom=geom_point(
                size=1.8,
            ),
            segment=geom_segment(
                linewidth = 0.3)
            )
        )
)

upsetplot

```

```{r}
# pdf("../plots/upset_plot.pdf", height = 5.5, width = 8)
# upsetplot
# dev.off()
```
