---
title: "Testing Hypotheses"
author: "Sofie Agerbaek"
date: "2026-01-29"
output: html_document
---

#### In this script I statistically test differences across samples at different taxonomic levels (species, genus, family). Generally I employ non-parametric tests.


```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(heatmaply)
library(writexl)
library(Polychrome)
library(tidyr)
library(stringr)
library(rstatix)
library(emmeans)
```

```{r, include=FALSE}
library(cowplot)
options(digits = 3)        # number of digits printed by R default (vectors, data.frames, lists)
options(pillar.sigfig = 3) # number of digits printed by tibbles default.
options(width = 200)

text_base_size   <- 9 # in pt
fig.witdh        <- 180  # in mm
fig.height       <- 125  # in mm

# Set all text in plots to same size
theme_set(theme_cowplot(font_size = 10, rel_small = 1, rel_tiny = 1, rel_large = 1))
theme_update(panel.grid.major = element_line(colour="gray85"), panel.grid.minor = element_line(colour="gray95", linetype = 7))
# Setting output sizes for plots
knitr::opts_chunk$set(fig.width = fig.witdh/25.4)
knitr::opts_chunk$set(fig.height = fig.height/25.4)
knitr::opts_chunk$set(dpi = 108) # You need to find your monitors dpi yourself.

# Setting text size inside plots (geom_text, geom_label etc.)
ggplot_text_size <- text_base_size / ggplot2::.pt
# Now use: geom_text(..., size = ggplot_text_size)

options(ggplot2.continuous.fill  = scale_fill_viridis_c)
options(ggplot2.discrete.fill    = c("slategray2", "#404e67","#809bce","#9887D9" ,"#9C5BA2", "#DB9FF6",  "#d1625c", "#ffd6a5", "#e89c81"))
options(ggplot2.continuous.colour = scale_colour_viridis_c)
options(ggplot2.discrete.colour   = c("slategray2","#404e67", "#809bce","#9C5BA2","#9887D9",  "#d1625c", "#ffd6a5", "#e89c81"))

# heatmap palette 
article_pallette <- c("#5674b4", "#7eacd1", "#afd9e9", "#e1f2f9", "#f7e18e", "#f7b060", "#ee6d43", "#d32e28")

```

##### Setup
Taxonomy dictionary
```{r}
taxonomy <- data.frame(
  sample = c(
    "Aspergillus_fumigatus_IP24",
    "Candida_albicans_F133_05",
    "Candida_dubliniensis_11945",
    "Candida_dubliniensis_2747",
    "Candida_dubliniensis_8501",
    "Candida_metapsilosis_Y_48470",
    "Candida_orthopsilosis_Y_48468",
    "Candida_parapsilosis_11301",
    "Candida_parapsilosis_1954",
    "Candida_tropicalis_MYCT_0816",
    "Chrysosporium_queenslandicum_28077",
    "Cryptococcus_neoformans_KN99",
    "Epidermophyton_floccosum_100148",
    "Epidermophyton_floccosum_10867",
    "Epidermophyton_floccosum_130802",
    "Histoplasma_capsulatum_G186A",
    "Hortaea_werneckii_100455",
    "Hortaea_werneckii_1163",
    "Hortaea_werneckii_41051",
    "Malassezia_dermatis_9169",
    "Malassezia_equina_9969",
    "Malassezia_furfur_141402",
    "Malassezia_furfur_1878",
    "Malassezia_furfur_6000",
    "Malassezia_globosa_7705",
    "Malassezia_globosa_7886",
    "Malassezia_globosa_8744",
    "Malassezia_pachydermatis_21004",
    "Malassezia_restricta_7877",
    "Malassezia_restricta_7991",
    "Malassezia_sympodialis_44340",
    "Microsporum_audouinii_119449",
    "Microsporum_audouinii_33268",
    "Microsporum_canis_113480",
    "Microsporum_canis_15669",
    "Microsporum_canis_44551",
    "Microsporum_ferrugineum_131111",
    "Microsporum_ferrugineum_37371",
    "Nakaseomyces_glabratus_11311",
    "Nakaseomyces_glabratus_138",
    "Nakaseomyces_glabratus_15698",
    "Nannizzia_gypsea_17164",
    "Pichia_kudriavzevii_5732",
    "Rhodotorula_toruloides_Z1",
    "Saccharomyces_cerevisiae_wine010",
    "Trichophyton_concentricum_109405",
    "Trichophyton_concentricum_19626",
    "Trichophyton_indotineae_BE17",
    "Trichophyton_interdigitale_BE4",
    "Trichophyton_kuryangei_IHEM13979",
    "Trichophyton_mentagrophytes_124404",
    "Trichophyton_mentagrophytes_43573",
    "Trichophyton_rubrum_117539",
    "Trichophyton_rubrum_17065",
    "Trichophyton_rubrum_28986",
    "Trichophyton_rubrum_73588",
    "Trichophyton_schoenleinii_45859",
    "Trichophyton_schoenleinii_85571",
    "Trichophyton_soudanese_IHEM19743",
    "Trichophyton_tonsurans_112188",
    "Trichophyton_tonsurans_12935",
    "Trichophyton_tonsurans_72988",
    "Trichophyton_violaceum_73088",
    "Trichosporon_ovoides_4098",
    "Trichosporon_ovoides_7612",
    "Trichosporon_ovoides_9430"
  ),
  species = c(
    "Aspergillus_fumigatus" ,"Candida_albicans","Candida_dubliniensis"
    ,"Candida_dubliniensis", "Candida_dubliniensis","Candida_metapsilosis",
    "Candida_orthopsilosis","Candida_parapsilosis", "Candida_parapsilosis","Candida_tropicalis","Chrysosporium_queenslandicum",
    "Cryptococcus_neoformans","Epidermophyton_floccosum","Epidermophyton_floccosum",
    "Epidermophyton_floccosum","Histoplasma_capsulatum","Hortaea_werneckii" ,"Hortaea_werneckii","Hortaea_werneckii","Malassezia_dermatis","Malassezia_equina",
    "Malassezia_furfur","Malassezia_furfur","Malassezia_furfur","Malassezia_globosa",
    "Malassezia_globosa","Malassezia_globosa","Malassezia_pachydermatis",
    "Malassezia_restricta","Malassezia_restricta","Malassezia_sympodialis",
    "Microsporum_audouinii","Microsporum_audouinii","Microsporum_canis",
    "Microsporum_canis","Microsporum_canis","Microsporum_ferrugineum",
    "Microsporum_ferrugineum","Nakaseomyces_glabratus","Nakaseomyces_glabratus",
    "Nakaseomyces_glabratus", "Nannizzia_gypsea","Pichia_kudriavzevii",
    "Rhodotorula_toruloides","Saccharomyces_cerevisiae", "Trichophyton_concentricum",
    "Trichophyton_concentricum" ,   "Trichophyton_indotineae"  ,   "Trichophyton_interdigitale",
    "Trichophyton_kuryangei"   ,    "Trichophyton_mentagrophytes" , "Trichophyton_mentagrophytes", 
    "Trichophyton_rubrum"      ,    "Trichophyton_rubrum"     ,    "Trichophyton_rubrum"    ,   
    "Trichophyton_rubrum"     ,     "Trichophyton_schoenleinii"  ,  "Trichophyton_schoenleinii",
    "Trichophyton_soudanese"     ,  "Trichophyton_tonsurans"   ,   "Trichophyton_tonsurans"  , 
    "Trichophyton_tonsurans"    ,   "Trichophyton_violaceum"  ,     "Trichosporon_ovoides"   ,    
    "Trichosporon_ovoides"       ,  "Trichosporon_ovoides"),
  genus = c(
    "Aspergillus",
    rep("Candida", 9),
    "Chrysosporium",
    "Cryptococcus",
    rep("Epidermophyton", 3),
    "Histoplasma",
    rep("Hortaea", 3),
    rep("Malassezia", 12),
    rep("Microsporum", 7),
    rep("Nakaseomyces", 3),
    "Nannizzia",
    "Pichia",
    "Rhodotorula",
    "Saccharomyces",
    rep("Trichophyton", 18),
    rep("Trichosporon", 3)
  ),
  family = c(
    "Trichocomaceae",     # Aspergillus
    rep("Debaryomycetaceae", 9),  # Candida spp.
    "Onygenaceae",        # Chrysosporium
    "Tremellaceae",       # Cryptococcus
    rep("Arthrodermataceae", 3),  # Epidermophyton
    "Ajellomycetaceae",   # Histoplasma
    rep("Teratosphaeriaceae", 3), # Hortaea
    rep("Malasseziaceae", 12),    # Malassezia spp.
    rep("Arthrodermataceae", 7),  # Microsporum spp.
    rep("Saccharomycetaceae", 3), # Nakaseomyces glabratus
    "Arthrodermataceae",  # Nannizzia
    "Saccharomycetaceae", # Pichia
    "Sporidiobolaceae",   # Rhodotorula
    "Saccharomycetaceae", # Saccharomyces
    rep("Arthrodermataceae", 18), # Trichophyton spp.
    rep("Trichosporonaceae", 3)   # Trichosporon spp.
  ),
  stringsAsFactors = FALSE
)

```

species order (for plotting)
```{r}
species_order <- c("Rhodotorula toruloides", "Cryptococcus neoformans", "Trichosporon ovoides",
                   "Malassezia furfur", "Malassezia equina", "Malassezia dermatis", "Malassezia sympodialis", 
                   "Malassezia pachydermatis", "Malassezia restricta", "Malassezia globosa", 
                   "Pichia kudriavzevii", "Saccharomyces cerevisiae", "Nakaseomyces glabratus",
                   "Candida metapsilosis", "Candida orthopsilosis", "Candida parapsilosis", "Candida tropicalis", 
                   "Candida albicans", "Candida dubliniensis",
                   "Hortaea werneckii",  "Aspergillus fumigatus", "Histoplasma capsulatum", 
                   "Chrysosporium queenslandicum", 
                   "Microsporum canis", "Microsporum audouinii", "Microsporum ferrugineum", 
                   "Nannizzia gypsea", "Epidermophyton floccosum", "Trichophyton schoenleinii",
                   "Trichophyton tonsurans", "Trichophyton mentagrophytes", "Trichophyton indotineae",
                   "Trichophyton interdigitale", "Trichophyton concentricum", "Trichophyton soudanese",
                   "Trichophyton kuryangei", "Trichophyton rubrum", "Trichophyton violaceum" )
```




## KEGG & EC

KEGG
```{r}
summary_df <- read.csv("../generated/species_kegg_category_summary.csv")
```

EC
```{r}
hydro_summary <- read.csv("../generated/EC_hydrolase_summary.csv")
```


Plotting KEGG
```{r}
selected_categories <- c("Amino acid metabolism", "Carbohydrate metabolism", "Energy metabolism", 
                         "Lipid metabolism", "Nucleotide metabolism")

plot_summary_df <- summary_df %>%
  group_by(species) %>%
  mutate(catprop = GeneCount / sum(GeneCount) * 100) %>%
  ungroup() %>%
  filter(Category %in% selected_categories) 

plot_summary_df$species <- sub("_", " ", plot_summary_df$species)
plot_summary_df$Category <- sub("metabolism", "", plot_summary_df$Category)

plot_summary_df$species <- factor(plot_summary_df$species, levels = rev(species_order))

KEGG_plot <- ggplot(plot_summary_df, aes(y = species, x = catprop, fill = Category)) +
  geom_bar(stat = "identity", position = "identity", alpha = 0.9) +  
  theme_minimal() +
  theme(axis.ticks.x = element_blank(), panel.spacing = unit(1.3, "lines"), axis.title = element_blank(),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = "grey80", linetype = "dashed"),
        axis.text.y = element_text(face = "bold.italic", color = "black", size = 7),
        strip.background = element_blank(),
        strip.text = element_text(size = 8, face = "bold", color = "black")) +
  facet_wrap(~Category, nrow = 1, scales = "free_x")+
#  expand_limits(x = c(NA, 15))+
  guides(fill = 'none')
KEGG_plot

```


### Testing differences in % of genes in functional categories

```{r}
taxonomy_subset <- taxonomy %>% dplyr::select(-sample) %>% mutate(species = sub("_", " ", species)) %>%
  group_by(species) %>% summarise(genus = first(genus), family = first(family))

KEGG_df <- left_join(plot_summary_df, taxonomy_subset, by = "species")

hydro_summary$species <- sub("_", " ", hydro_summary$species)
EC_df <- left_join(hydro_summary, taxonomy_subset, by = "species")
```

Kruskal–Wallis test, is at least one family different in each category

```{r}
print(unique(KEGG_df$genus))
print(unique(KEGG_df$family))

kw_results <- KEGG_df %>%
  group_by(Category) %>%
  kruskal_test(catprop ~ genus) %>%
  mutate(p.adj = p.adjust(p, method = "BH"))

kw_results

kw_results <- KEGG_df %>%
  group_by(Category) %>%
  kruskal_test(catprop ~ family) %>%
  mutate(p.adj = p.adjust(p, method = "BH"))

kw_results


## EC GENERA

kw_results <- EC_df %>%
  group_by(Hydrolase_sub) %>%
  kruskal_test(catprop ~ genus) %>%
  mutate(p.adj = p.adjust(p, method = "BH"))

kw_results

## EC FAMILY

kw_results <- EC_df %>%
  group_by(Hydrolase_sub) %>%
  kruskal_test(catprop ~ family) %>%
  mutate(p.adj = p.adjust(p, method = "BH"))

kw_results

```

KEGG pairwise
```{r}
#For KEGG all except carbohydrates were significant 

KEGG_filt <- KEGG_df %>% filter(Category != "Carbohydrate ") 

print(unique(KEGG_filt$genus))

posthoc <- KEGG_filt %>%
  group_by(Category) %>%
  dunn_test(catprop ~ genus, p.adjust.method = "BH")

posthoc[posthoc$p.adj.signif != "ns",]

#write.csv(posthoc, "../generated/KEGG_pw_genus.csv")

KEGG_filt <- KEGG_df %>% filter(Category != "Carbohydrate ") 

print(unique(KEGG_filt$family))

posthoc <- KEGG_filt %>%
  group_by(Category) %>%
  dunn_test(catprop ~ family, p.adjust.method = "BH")

posthoc[posthoc$p.adj.signif != "ns",]

#write.csv(posthoc, "../generated/KEGG_pw_family.csv")
```


EC pairwise
```{r}
# all EC categories were significant in the KW test

# Genus
EC_filt <- EC_df 

posthoc <- EC_filt %>%
  group_by(Hydrolase_sub) %>%
  dunn_test(catprop ~ genus, p.adjust.method = "BH")

posthoc[posthoc$p.adj.signif != "ns",]

#write.csv(posthoc, "../generated/EC_pw_genus.csv")

# Family

posthoc <- EC_filt %>%
  group_by(Hydrolase_sub) %>%
  dunn_test(catprop ~ family, p.adjust.method = "BH")

posthoc[posthoc$p.adj.signif != "ns",]

#write.csv(posthoc, "../generated/EC_pw_family.csv")
```

```{r}
rm(EC_df, EC_filt, hydro_summary, KEGG_df, KEGG_filt, posthoc, kw_results, plot_summary_df, summary_df, selected_categories, KEGG_plot)
```


## Predicted secreted metabolic enzymes - Fig 5 


### Number of predicted secreted across all - testing w binomial dist if they're significantly different

```{r}
all_anno <- read.csv("../generated/all_annotations_combined.tsv")
all_anno$X <- NULL
```


```{r}
secreted <- all_anno %>% filter(signalp == "True")

sum(!is.na(all_anno$signalp))

secr_sum <- all_anno %>% group_by(strain_id) %>% summarise(species = first(species), tot_genes = n(), secr = sum(!is.na(signalp))) %>% 
            mutate(prop_secr <- secr/tot_genes)


secr_sum$genus <- str_split_fixed(secr_sum$species, "_", 2)[, 1]
```

Binomial testing (its two-case scenario, but the number of gene difference also matter) - quasi handles the biased sample sizes better
```{r}
model <- glm(cbind(secr, tot_genes - secr) ~ species,
             data = secr_sum, 
             family = quasibinomial)

anova(model, test = "F")
# Significant effect of species on the proportion of secreted genes
```

pairwise
```{r}
pw_secr_test <- as.data.frame(emmeans(model, pairwise ~ species, type = "response", adjust = "BH")$contrasts)
pw_secr_test$sig <- pw_secr_test$p.value < 0.05

pw_secr_test$sample1 <- str_split_fixed(pw_secr_test$contrast, " / ", 2)[, 1]
pw_secr_test$sample2 <- str_split_fixed(pw_secr_test$contrast, " / ", 2)[, 2]

t <- pw_secr_test %>% filter(sig == "TRUE") 
head(t)
table(rbind(t$sample1, t$sample2))

#write.csv(pw_secr_test, "../generated/pw_secreted_test.csv")
```


### Secreted enzymes with different targets

#### Setup
```{r}
library(readxl)
test_cur <- read_xlsx("../generated/1211_manual_cur_secreted.xlsx")
test_cur$netgpi_geneswithgpi <- as.numeric(test_cur$netgpi_geneswithgpi)
```

```{r}
collapsed_df <- test_cur %>% 
  filter(!category == "NA") %>%
  dplyr::select(-c("description", "IPR", "PFAM", "EC", "COG", "canonical", "pfam_names")) %>%
  mutate(desc_collapse = str_trim(tolower(desc_collapse))) %>%
  group_by(desc_collapse) %>%
  summarise(OG = paste(unique(OG), collapse = "; "),
            category = first(category),
            number_of_species = max(number_of_species, na.rm = TRUE),
            total_genec = sum(total_genec, na.rm = TRUE),
            number_of_genera = max(number_of_genera, na.rm = TRUE),
            number_of_families = max(number_of_families, na.rm = TRUE),
            signalp = sum(as.numeric(signalp), na.rm = TRUE),
            tmhmm_geneswtmh = sum(as.numeric(tmhmm_geneswtmh), na.rm = TRUE),
            netgpi_geneswithgpi = sum(netgpi_geneswithgpi, na.rm = TRUE),
            across(all_of(names(.)[9:(ncol(.)-1)]), ~ sum(.x, na.rm = TRUE)),
            )

```

```{r}
collapsed_filtered <- collapsed_df %>%
  filter(tmhmm_geneswtmh < 0.7*signalp) %>%
  filter(number_of_species > 2)

plot_df <- collapsed_filtered %>% dplyr::select(desc_collapse, category, names(collapsed_df)[11:76])

plot_df_long <- pivot_longer(plot_df, cols = -c(desc_collapse, category), names_to = "sample", values_to = "gene_copies")

plot_df_long <- merge(plot_df_long, taxonomy[1:2], by = "sample")
length(unique(plot_df_long$sample))
```


Since we consider all genes in an OG as secreted if more than 1/3 of the genes are signalp positive and less than 70% TMHMM positive, we should consider the offset to be the sum of all genes (in a sample) that are in OGs considered secreted.
```{r}
total_secr_OG_genec_sampl <- colSums(test_cur[15:80])

test_secr_df <- plot_df_long 
test_secr_df$total_secreted <- total_secr_OG_genec_sampl[test_secr_df$sample]

subset_cats <- c("carbohydrate metabolism", "lipid metabolism", "protein metabolism")

test_secr_df <- test_secr_df %>% filter(category %in% subset_cats)
test_secr_df$sample <- sub("-", "_", test_secr_df$sample)

test_secr_df <- left_join(test_secr_df, taxonomy, by = "sample")

#tryout sans hybrids
# hybrid_samples <- c("Malassezia_furfur_1878", "Malassezia_furfur_6000", "Trichosporon_ovoides_9430", "Hortaea_werneckii_41051", "Hortaea_werneckii_100455")
# test_secr_df <- test_secr_df %>% filter(!sample %in% hybrid_samples)

```


Plotting 


plotting the metabolism
```{r, fig.show="inline", fig.width=10, fig.height=8}
library(pheatmap)

plot_df <- read.csv("../generated/plot5_df.csv", check.names=FALSE, row.names = 1)

subset_cats <- c("carbohydrate metabolism", "lipid metabolism", "protein metabolism")

# Filter and order the dataframe
plot_df_sub <- plot_df %>%
  filter(category %in% subset_cats) %>%
  mutate(category = factor(category, levels = subset_cats)) %>%
  arrange(category)

# numeric matrix
num_df <- plot_df_sub[, species_order] %>%
  mutate(across(everything(), as.numeric))
mat <- as.matrix(num_df)
rownames(mat) <- rownames(plot_df_sub)

# Z scores for colour scale
mat_z <- t(scale(t(mat)))  # center and scale each row
rownames(mat_z) <- rownames(mat)
mat_z[mat_z > 2] <- 2
mat_z[mat_z < -2] <- -2

# ----- 4. Build annotation -----
annotation_row <- data.frame(Category = plot_df_sub$category)
rownames(annotation_row) <- rownames(mat)  # align annotation with matrix

# ----- 5. Compute white gaps -----
gaps_row <- cumsum(table(plot_df_sub$category))
gaps_row <- gaps_row[1:3]

# ---- Numeric labels ----
num_labels <- matrix(round(mat, 1), nrow = nrow(mat), ncol = ncol(mat))
rownames(num_labels) <- rownames(mat)
colnames(num_labels) <- colnames(mat)

# ---- Plot ----
plot <- pheatmap(mat_z,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         #annotation_row = annotation_row,
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_row = 10,        # size of y-axis (row) labels
         fontsize_col = 12,       # size of x-axis labels
         display_numbers = num_labels,
         number_color = "black",
         fontsize_number = 8,
         gaps_row = gaps_row,
         border_color = "white",
         color = article_pallette, 
         angle_col = 90,
         legend = FALSE,
         cellheight=9,
         cellwidth=10)

plot
rownames(mat_z)
```


#### Carbohydrates 

```{r, include=FALSE}
library(MASS)
```

```{r, warning=FALSE}
df_carbs <- test_secr_df %>% filter(category == "carbohydrate metabolism")

model_carbs <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
       data = df_carbs)

anova(model_carbs, test = "Chisq")

```

Testing each narrow category by iterating over them
```{r, warning=FALSE}

res_carb <- df_carbs %>%
  group_by(desc_collapse) %>%
  group_modify(~{
    m <- glm.nb(
      gene_copies ~ genus + offset(log(total_secreted)),
      data = .x
    )
    a <- anova(m, test = "Chisq")
    tibble(
      p_value = a$`Pr(>Chi)`[2]
    )
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"), sig = p_adj<0.05) 

print(res_carb[res_carb$sig == FALSE,])
non_sig_cats <- res_carb[res_carb$sig == FALSE,]$desc_collapse
#Chitosanase is not sig, we take it out of the pairwise

lst_carb <- list()

df_carbs %>%
  filter(!desc_collapse %in% non_sig_cats) %>%
  group_by(desc_collapse) %>%
  group_walk(~{
    # .x is the data for this category, .y$desc_collapse is the category name
    m <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
                data = .x)
    
    # pairwise contrasts
    pw <- as.data.frame(emmeans(m, pairwise ~ genus, type="response", adjust="BH")$contrasts)
    
    # store in the list under the category name
    lst_carb[[.y$desc_collapse]] <<- pw  # note the <<- for global assignment
  })

head(lst_carb$`sugar metabolism`)
```

Save individual pairwise pr. annotation as one long table 
```{r}
long_carb <- bind_rows(lst_carb, .id = "annotation")
#write.csv(long_carb, "../generated/secr_carb_pw.csv")
rm(long_carb, lst_carb)
```


#### Lipid

```{r, warning=FALSE}
df_lipid <- test_secr_df %>% filter(category == "lipid metabolism") 

model_lipid <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
       data = df_lipid)

anova(model_lipid, test = "Chisq")
```


```{r, warning=FALSE}
res_lipid <- df_lipid %>%
  group_by(desc_collapse) %>%
  group_modify(~{
    m <- glm.nb(
      gene_copies ~ genus + offset(log(total_secreted)),
      data = .x
    )
    a <- anova(m, test = "Chisq")
    tibble(
      p_value = a$`Pr(>Chi)`[2]
    )
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"), sig = p_adj<0.05) 

print(res_lipid[res_lipid$sig == FALSE,])
non_sig_cats <- res_lipid[res_lipid$sig == FALSE,]$desc_collapse
lst_lipid <- list()

df_lipid %>%
  filter(!desc_collapse %in% non_sig_cats) %>%
  group_by(desc_collapse) %>%
  group_walk(~{
    # .x is the data for this category, .y$desc_collapse is the category name
    m <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
                data = .x)
    
    # pairwise contrasts
    pw <- as.data.frame(emmeans(m, pairwise ~ genus, type="response", adjust="BH")$contrasts)
    
    # store in the list under the category name
    lst_lipid[[.y$desc_collapse]] <<- pw  # note the <<- for global assignment
  })

head(lst_lipid$lipase)
```

Save individual pairwise pr. annotation as one long table 
```{r}
long_lipid <- bind_rows(lst_lipid, .id = "annotation")
#write.csv(long_lipid, "../generated/secr_lipid_pw.csv")
rm(long_lipid, lst_lipid)
```


#### Protein


```{r, warning=FALSE}
df_protein <- test_secr_df %>% filter(category == "protein metabolism")

model_protein <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
       data = df_protein)

anova(model_protein, test = "Chisq")

```

```{r, warning=FALSE}
res_protein <- df_protein %>%
  group_by(desc_collapse) %>%
  group_modify(~{
    m <- glm.nb(
      gene_copies ~ genus + offset(log(total_secreted)),
      data = .x
    )
    a <- anova(m, test = "Chisq")
    tibble(
      p_value = a$`Pr(>Chi)`[2]
    )
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"), sig = p_adj<0.05) 

print(res_protein[res_protein$sig == FALSE,])
non_sig_cats <- res_protein[res_protein$sig == FALSE,]$desc_collapse

lst_protein <- list()

df_protein %>%
  filter(!desc_collapse %in% non_sig_cats) %>%
  group_by(desc_collapse) %>%
  group_walk(~{
    # .x is the data for this category, .y$desc_collapse is the category name
    m <- glm.nb(gene_copies ~ genus + offset(log(total_secreted)),
                data = .x)
    
    # pairwise contrasts
    pw <- as.data.frame(emmeans(m, pairwise ~ genus, type="response", adjust="BH")$contrasts)
    
    # store in the list under the category name
    lst_protein[[.y$desc_collapse]] <<- pw  # note the <<- for global assignment
  })

head(lst_protein$`aspartyl protease`)
```

Save individual pairwise pr. annotation as one long table 
```{r}
long_protein <- bind_rows(lst_protein, .id = "annotation")
#rite.csv(long_protein, "../generated/secr_protein_pw.csv")
rm(long_protein, lst_protein)
```


Save all res tables in one file
```{r}
full_res <- rbind(res_carb, res_lipid, res_protein)
#write.csv(full_res, "../generated/secreted_cat_anova.csv")
```

```{r, include=FALSE, warning=FALSE}
rm(df_carbs, df_lipid, df_protein, glm_rep, mat, mat_z, model, model_carbs, model_lipid, model_protein, pairwise_protein, pw, pw_secr_test, res_carb, res_lipid, res_protein, secreted, test_cur, test_secr_df, plot, plot_df, plot_df_long, plot_df_sub, secr_sum, t, taxonomy_subset, num_df, collapsed_df, collapsed_filtered, num_labels, annotation_row)
```


## Adhesins 

```{r}
total_genes <- all_anno %>% group_by(strain_id) %>% summarise(species = first(species), tot_genes = n())

# All_anno has old strain numbers, they need to be updated:
old_nr_to_cbs <- c(
  "1" = "10867",
  "3" = "4098", 
  "6" = "112188",
  "8" = "45859",
  "9" = "124404",
  "10" = "1163",
  "11" = "138",
  "12" = "1954",
  "13" = "15669",
  "27" = "7705",
  "28" = "7877",
  "29" = "141402",
  "31" = "5732",
  "32" = "28986",
  "34" = "2747",
  "410" = "41051",
  "332" = "33268",
  "445" = "44551",
  "171" = "17164", 
  "435" = "43573",
  "735" = "73588",
  "855" = "85571",
  "129" = "12935",
  "739" = "72988",
  "730" = "73088")

total_genes <- total_genes %>%
  mutate(strain_id = if_else(strain_id %in% names(old_nr_to_cbs),
                             old_nr_to_cbs[strain_id],
                             strain_id))

total_genes$sample <- paste(total_genes$species, total_genes$strain_id, sep = "_")
```


```{r}
gpi_plot_df_long <- read.csv("../generated/gpi_plot_df_long.csv")

gpi_sum <- gpi_plot_df_long %>% filter(category == "adhesin-associated") %>% group_by(sample) %>% summarise(tot_gpi = sum(gene_copies))

gpi_sum <- left_join(gpi_sum, total_genes, by = 'sample')

gpi_sum$non_gpi_genes <- gpi_sum$tot_genes - gpi_sum$tot_gpi

gpi_sum$sample <- sub("-", "_", gpi_sum$sample)
gpi_sum <- left_join(gpi_sum, taxonomy, by = c("sample", "species"))

gpi_sum <- gpi_sum %>%
  mutate(
    genus = as.factor(genus),
    species = as.factor(species)
  )

gpi_sum$adhesin_prop <- gpi_sum$tot_gpi / gpi_sum$tot_genes
```

Kruskal-Wallis test - is any across all significantly different? 

```{r}
kruskal.test(adhesin_prop ~ genus, data = gpi_sum)
#significant
```

Doing all vs. rest to test which genera (if any) are significantly different than the rest:
```{r}
res <- lapply(levels(gpi_sum$genus), function(g) {
  w <- wilcox.test(adhesin_prop ~ (genus == g), data = gpi_sum, p.adjust.method = "BH")
  data.frame(genus = g, p_value = w$p.value)
})

one_vs_rest_genus <- do.call(rbind, res)
one_vs_rest_genus$p_adj <- p.adjust(one_vs_rest_genus$p_value, method = "BH")

one_vs_rest_genus[order(one_vs_rest_genus$p_adj),]
```

